---
title: "2020 SISBID Graphical Models Lab"
author: "Genevera I. Allen, Yufeng Liu, Hui Shen, Camille Little"
date: "7/21/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r message= FALSE, warning= FALSE}
library("igraph")
library("XMRF")
library("huge")
library("glasso")
library("WGCNA")
library("glmnet")
```

Read the Sachs et al data: Flow cytometry proteomics in single cells, $p = 11$ proteins measured in $n = 6466$ cells.
```{r}
sachscov <- as.matrix(read.table("sachscov.txt"))
sachscor <- cov2cor(sachscov)
sachsdat <- as.matrix(read.table("sachs.data.txt"))
ps <- c("praf","pmek","plcg","PIP2","PIP3","P44","pakts","PKA","PKC","P38","pjnk")
colnames(sachsdat) <- ps
p <- ncol(sachsdat)
n <- nrow(sachsdat)
dim(sachsdat)
dim(sachscov)
head(sachsdat)
```


```{r}
load("UnsupL_SISBID_2020.Rdata")
```

### Coexpression network 
#### simple thresholding of correlations, at a cutoff chosen to give similar number of edges to partial correlation methods  
* a randomly chosen threshold
```{r}
tau <- 0.1
A1 <- abs(sachscor) > tau
diag(A1) <- 0
sum(A1)/2
```


#### testing for nonzero correlations     
* testing for nonzero correlation, using Fisher Z-transform
```{r}
fisherzs <- atanh(sachscor)
fisherps  <- 2*pnorm(abs(fisherzs), 0, 1/sqrt(n-3), lower.tail=FALSE)
A2 <- fisherps < (0.01/(p*(p-1)/2))
diag(A2) <- 0
sum(A2)/2
```

## plot the three networks
```{r}
g1 <- graph.adjacency(A1, mode="undirected")
g2 <- graph.adjacency(A2, mode="undirected")

g0 <- g2
```

```{r}
plot(g1,layout=layout.circle(g1), main='simple thresholding of correlations')
plot(g2,layout=layout.circle(g2), main='testing for nonzero correlations')
```

## Partial correlation networks(under gaussian graphical models assumption)  
inverse covariance matrix
```{r}
invcov <- abs(round(solve(sachscor),3)) 
invcor <- cov2cor(invcov)
A1 <- 1*(invcor > 0.05)
diag(A1) <- 0
sum(A1)/2
g1 <- graph.adjacency(A1, mode="undirected")
```

```{r}
plot(g1,layout=layout.circle(g1),main = "Partial correlation networks")
```

Graphical lasso  
Calculate lambda, based on formula in the slides (the third method)
```{r}
alpha = 0.01
num = qt(p=alpha/(2*(p^2)),df=n-2, lower.tail=F)
lambda = num / sqrt(n-2 + num)
```

Apply glasso
```{r}
glasso.est <- glasso(s=sachscor,rho=lambda*4.2,approx=FALSE,
                     penalize.diagonal=FALSE)
A2 <- abs(glasso.est$wi) > 1E-16
diag(A2) <- 0
g2 <- graph.adjacency(A2, mode="undirected")
```

Neighborhood selection
```{r}
ns.est <- glasso(s=sachscor, rho=lambda, approx=TRUE, penalize.diagonal=FALSE)
A3 <- abs(ns.est$wi) > 1E-16; diag(A3) <- 0
g3 <- graph.adjacency(A3, mode="undirected")
```

Neighborhood selection estimate with huge (Stability selection for the value of $\lambda$)  
```{r}
X <- data.matrix(scale(sachsdat))
neth = huge(X,method="mb")
plot(neth)
```


```{r}
## stability selection with huge
net.s <- huge.select(neth, criterion="stars")
plot(net.s)
```


```{r}
#larger lambda
mat = neth$path[[2]]
neti = as.undirected(graph_from_adjacency_matrix(mat))
plot(neti,vertex.label=colnames(X),vertex.size=2,vertex.label.cex=1.2,vertex.label.dist=1,layout=layout_with_kk)
```


```{r}
#smaller lambda
mat = neth$path[[5]]
neti = as.undirected(graph_from_adjacency_matrix(mat))
plot(neti,vertex.label=colnames(X),vertex.size=2,vertex.label.cex=1.2,vertex.label.dist=1,layout=layout_with_kk)
```

## Nonparanormal Models: rank-based correlation
```{r}
scor <- cor(sachsdat,method='spearman')
scor <- 2*sin(scor*pi/6)
npn.est <- glasso(s=scor, rho=lambda, approx=FALSE, penalize.diagonal=FALSE)
A4 <- abs(npn.est$wi) > 1E-16
diag(A4) <- 0
g4 <- graph.adjacency(A4, mode="undirected")
```

## Nonparanormal Models -- alternative estiamtion
```{r}
npn.cor <- huge.npn(x=sachsdat, npn.func="skeptic", npn.thresh=NULL, verbose=FALSE)
npn.est <- glasso(s=npn.cor, rho=lambda, penalize.diagonal=FALSE)
A5 <- abs(npn.est$wi) > 1E-16
diag(A5) <- 0
g5 <- graph.adjacency(A5, mode="undirected")
```

## binary network estimation
```{r}
sachsbin <- 1*(sachsdat > 0) + -1*(sachsdat <= 0)
head(sachsbin)
bin.est <- matrix(0,p,p)
## estiamte the neighborhood for each node 
for(j in 1:p){
  ## this is the same method used in neighborhood selection, the only difference is 'family'
  nbr <- glmnet(x=sachsbin[,-j], y=sachsbin[,j], family='binomial', lambda=lambda) 
  bin.est[j,-j] <- 1*(abs(as(nbr$beta,"matrix")) > 0)	#store the estimates in jth row of matrix
}
A6 <- bin.est
diag(A6) <- 0
sum(A6)/2
g6 <- graph.adjacency(A6, mode="undirected")
```

## plot the networks
```{r}
plot(g1,layout=layout.circle(g1), main='Partial correlation networks')
plot(g2,layout=layout.circle(g2), main='Glasso')
plot(g3,layout=layout.circle(g3), main='Neighborhood selection')
plot(g4,layout=layout.circle(g4), main='nonparanormal')
plot(g5,layout=layout.circle(g5), main='nonparanormal - v2')
plot(g6,layout=layout.circle(g6), main='Binary')
```


#### Poisson Graphical Models via XMRF package  
Data set - Author Data. This data set consists of word counts from chapters written by four British authors.  
```{r}
TrueAuth = as.factor(rownames(author))
X2 = log(1 + author[,1:69]) #log transform data
```

```{r}
lam = lambdaMax(X2)*sqrt(log(ncol(X2))/nrow(X2))*0.02
net = XMRF(t(X2),method="LPGM",lambda.path=lam,N=1,th=.001)
mat = net$network[[1]]
neti = as.undirected(graph_from_adjacency_matrix(mat))
plot(neti,vertex.label=colnames(X2),vertex.size=2,vertex.label.cex=1.2,vertex.label.dist=1,layout=layout_with_kk)

plot(neti,vertex.label=colnames(X2),vertex.size=2,vertex.label.cex=1.2,vertex.label.dist=1,layout=layout_in_circle)
```




