---
title: "2020 SISBID Dimension Reduction Lab"
author: "Genevera I. Allen,  Yufeng Liu, Hui Shen, Camille Little"
date: "7/20/2020"
output: pdf_document
---


# Quick PCA Demo Using College Data

Load in Packages
```{r, echo = TRUE}
library(ISLR)
library(ggplot2)
library(GGally)
```

Load Digits Data
```{r, echo = TRUE}
#code for digits - ALL
rm(list=ls())
load("UnsupL_SISBID_2020.Rdata")
```


```{r, echo  = TRUE}
data(College)
cdat = College[,2:18]
dim(cdat)
names(cdat)
```

```{r, echo = TRUE}
pc = princomp(cdat) #default - centers and scales

#Go back and display these plots side by side

biplot(pc,cex=.7)
screeplot(pc)

```

scatter plots - patterns among observations
```{r, echo = TRUE}
PC1 <- as.matrix(x=pc$scores[,1]) 
PC2 <- as.matrix(pc$scores[,2])

PC <- data.frame(State = row.names(cdat), PC1, PC2)
ggplot(PC, aes(PC1, PC2)) + 
  geom_text(aes(label = State), size = 3) +
  xlab("PC1") + 
  ylab("PC2") + 
  ggtitle("First Two Principal Components of College Data")

```

Pairs Plot
```{r, echo = TRUE}
comp_labels<-c("PC1","PC2","PC3","PC4", "PC5")
pairs(pc$scores[,1:5], labels = comp_labels, main = "Pairs of PC's for College Data")
```




Loadings - variables that contribute to these patterns
```{r, echo = TRUE}

par(mfrow=c(2,1))
barplot(pc$loadings[,1],cex.names=.6,main="PC 1 Loadings")
barplot(pc$loadings[,2],cex.names=.6,main="PC 2 Loadings")
```

Variance explained
```{r, echo = TRUE}

varex = 100*pc$sdev^2/sum(pc$sdev^2)
par(mfrow=c(2,1))
screeplot(pc)
plot(varex,type="l",ylab="% Variance Explained",xlab="Component")
```


Cumulative variance explained
```{r, echo = TRUE}
#cumulative variance explained
cvarex = NULL
for(i in 1:ncol(cdat)){
  cvarex[i] = sum(varex[1:i])
}
plot(cvarex,type="l",ylab="Cumulative Variance Explained",xlab="Component", main = "Principal Component V. Variance Explained" )
```

# Sparse PCA

```{r}
library(PMA)

spc = SPC(scale(cdat),sumabsv=2,K=3)

spcL = spc$v
rownames(spcL) = names(cdat)


```

Scatterplots of Sparse PCs
```{r, echo = TRUE}
i = 1; j = 2;
plot(spc$u[,i],spc$u[,j],pch=16,cex=.2, xlab = "PC 1", ylab = "PC 2", main = "Scatterplot of Sparse PC's ")
text(spc$u[,i],spc$u[,j],rownames(cdat),cex=.6)

```



Loadings
```{r, echo = TRUE}

par(mfrow=c(2,1))
barplot(spc$v[,1],names=names(cdat),cex.names=.6,main="SPC 1 Loadings")
barplot(spc$v[,2],names=names(cdat),cex.names=.6,main="SPC 2 Loadings")

```


# Try Princomp Function for Digits 3 and 8
```{r, echo = TRUE}
dat38 = rbind(digits[which(rownames(digits)==3),],digits[which(rownames(digits)==8),])
```

```{r, echo = TRUE}
pc = princomp(dat38) #default - centers and scales
```

Pairs plot Using ggpairs
```{r, echo = TRUE}
PC1 <- as.matrix(x=pc$scores[,1]) 
PC2 <- as.matrix(pc$scores[,2])
PC3 <- as.matrix(pc$scores[,3])
PC4 <- as.matrix(pc$scores[,4])
PC5<-as.matrix(pc$scores[,5])

pc.df.digits <- data.frame(digit_name = row.names(dat38), PC1, PC2,PC3, PC4, PC5)

ggpairs(pc.df.digits, mapping = aes(color = digit_name))
```


PC Loadings
```{r, echo = TRUE}
par(mfrow=c(3,5),mar=c(.1,.1,.1,.1))
for(i in 1:15){
  imagedigit(pc$loadings[,i])
}
```


# PCA LAB Using Digits Data

Data set - Digits Data.
Either use all digits or choose 2-3 digits if computational speed is a problem. Looking at 3's, 8's and 5's are interesting. 

## Problem 1 - PCA

Problem 1a - Apply PCA to this data.

Problem 1b - Do the first several PCs well separate different digits?  Why or why not?

Problem 1c - Use the first several PCs and PC loadings to evaluate the major patterns in the digits data.  Can you come up with a description of the pattern found by each of the first five PCs?

Problem 1d - How many PCs are needed to explain 95% of the variance?  You must decide how many PCs to retain.  Which do you pick and why?


## Problem 2 - MDS

Problem 2a - Apply MDS (classical or non-metric) to this data.  Try out several distance metrics and different numbers of MDS components.

Problem 2b - Which distance metric is best for this data?  Which one reveals the most separation of the digits?

Problem 2c - Compare and contrast the MDS component maps to the dimension reduction of PCA.  Which is preferable?

## Problem 3 - ICA.

Problem 3a - Apply ICA to this data set.

Problem 3b - Which value of K did you use?  Why?  What happens when you slightly change your chosen K?

Problem 3c - Interpret the independent image signals found.  Do any other them accurately reflect the different digits?  Which ones?

## Problem 4 - UMAP
Problem 5a - Apply UMAP on this data set. 

## Problem 5 - tSNE
Problem 5a - Apply tSNE on this data set 

## Problem 6 - Comparisons.
Problem 6a - Compare and contrast PCA, MDS and ICA, TSNE, and UMAP on this data set.  Which one best separates the different digits?  Which one reveals the most interesting patterns?

Problem 6b - Overall, which method do you recommend for this data set and why?


## Additional Data set - NCI Microarray data
(If you have time - take a further look at this data set using various methods for dimension reduction.  Also you may be interested in trying MDS to visualize this data.)


##R scripts to help out with the Dimension Reduction Lab
#Don't peek at this if you want to practice coding on your own!!

```{r}
library(ISLR)
library(ggplot2)
library(tidyr)
```


Load in data and visualize
```{r}
#code for digits - ALL
rm(list=ls())
load("UnsupL_SISBID_2020.Rdata")

#visulaize
pdf("temp.pdf")
par(mfrow=c(4,8), mar=c(.1,.1,.1,.1))
for(i in 1:32){
  imagedigit(digits[i,])
}
dev.off()

```

##Problem 1 - PCA

PCA - take SVD to get solution
don't center and scale to retain interpretation as images
```{r, echo = TRUE}
########Problem 1 - PCA
#PCA - take SVD to get solution
#don't center and scale to retain interpretation as images
svdd = svd(digits)
U = svdd$u
V = svdd$v #PC loadings
D = svdd$d
Z = digits%*%V #PCs


```

PC scatterplot
```{r, echo = TRUE}

i = 1; j = 3;
par(mfrow=c(1,1), mar=c(3,3,1,1))
plot(U[,i],U[,j],type="n", xlab = "PC1", ylab = "PC2", main = "PC Scatterplot of Digits")
text(U[,i],U[,j],rownames(digits),col=rownames(digits),cex=.7)


```

PC loadings
```{r, echo = TRUE}
#PC loadings
par(mfrow=c(3,5),mar=c(.1,.1,.1,.1))
for(i in 1:15){
  imagedigit(V[,i])
}

```



Variance Explained
```{r, echo = TRUE}

#Variance Explained
varex = 0
cumvar = 0
denom = sum(D^2)
for(i in 1:256){
  varex[i] = D[i]^2/denom
  cumvar[i] = sum(D[1:i]^2)/denom
}

```


Screeplot
```{r, echo = TRUE}

par(mfrow=c(1,2))
plot(1:256,varex,type="l",lwd=2,xlab="PC",ylab="% Variance Explained")
plot(1:256,cumvar,type="l",lwd=2,xlab="PC",ylab="Cummulative Variance Explained")
```

Pairs Plot
```{r, echo = TRUE}
library(GGally)
```

```{r, echo = TRUE}
Z_sub = Z[,1:5]
comp_labels<-c("PC1","PC2","PC3","PC4", "PC5")
pairs(Z_sub, labels = comp_labels, main = "Pairs of PC's for Digits Data")
```


# Problem 2 - MDS

classical MDS
(Note, this may take some time - try only on 3's and 8's)
```{r, echo = TRUE}

dat38 = rbind(digits[which(rownames(digits)==3),],digits[which(rownames(digits)==8),])
dim(dat38)

#PCA for comparison
svdd = svd(dat38)
U = svdd$u
V = svdd$v #PC loadings
D = svdd$d
Z = digits%*%V #PCs


```

```{r, echo = TRUE}
#MDS
Dmat = dist(dat38,method="maximum") #Manhattan (L1) Distance
mdsres = cmdscale(Dmat,k=10)

i = 1; j = 2;
par(mfrow=c(1,2))
plot(mdsres[,i],mdsres[,j],type="n", main = "MDS Using Manhattan Distance")
text(mdsres[,i],mdsres[,j],rownames(dat38),col=rownames(dat38))
```


```{r, echo = TRUE}
plot(U[,i],U[,j],type="n",xlab="PC1",ylab="PC2")
text(U[,i],U[,j],rownames(dat38),col=rownames(dat38))
```



# Problem 3 - ICA

```{r, echo = TRUE}
library(fastICA)
require("fastICA")

K = 20
icafit = fastICA(t(dat38),n.comp=K)

#plot independent source signals 
options(width = 60)
par(mfrow=c(4,5),mar = c(2, 2, 2, 2))
for(i in 1:K){
  imagedigit(icafit$S[,i])
}
```

# Problem 4 - UMAP

Install Packages
```{r, echo = TRUE}
#install.packages('umap')
#install.packages('Rtsne')
library(umap)
library(Rtsne)
```

Run UMAP
```{r, echo = TRUE}
digits.umap = umap(dat38)
```

Plot UMAP
```{r, echo = TRUE}

plot(digits.umap$layout[,1],y=digits.umap$layout[,2], type ='n', main = "UMAP on Digits 3,8 ", xlab = "UMAP1", ylab = "UMAP2")
text(digits.umap$layout[,1],y=digits.umap$layout[,2],rownames(dat38),col=rownames(dat38),cex=.7)

```

# Problem 5 - tSNE
Run tSNE
```{r, echo = TRUE}
tsne_digit <- Rtsne(as.matrix(dat38))
```

```{r, echo = TRUE}
plot(tsne_digit$Y[,1],y=tsne_digit$Y[,2], type ='n', main = "tSNE on Digits 3,8 ", xlab = "tSNE1", ylab = "tSNE2")
text(tsne_digit$Y[,1],y=tsne_digit$Y[,2],rownames(dat38),col=rownames(dat38),cex=.7)

```

