---
title: "2026 SISBID PCA Demo"
author: "Genevera I. Allen & Yufeng Liu"
output:
  pdf_document: default
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---


# Setup

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)

needed_packages <- c("ggplot2", "GGally")
missing_packages <- needed_packages[!vapply(needed_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop("Please install these packages before knitting: ", paste(missing_packages, collapse = ", "))
}

library(ggplot2)
library(GGally)

theme_set(theme_minimal(base_size = 12))
set.seed(1)
```

Helper functions used throughout the demo:

```{r helpers, echo = TRUE}
variance_table <- function(pca_object) {
  variance <- pca_object$sdev^2
  pct <- 100 * variance / sum(variance)
  data.frame(
    component = seq_along(pct),
    pct_variance = pct,
    cumulative_pct_variance = cumsum(pct)
  )
}

plot_variance <- function(var_tbl, title = "Variance explained", max_components = 30) {
  keep <- seq_len(min(max_components, nrow(var_tbl)))
  plot_df <- rbind(
    data.frame(
      component = var_tbl$component[keep],
      value = var_tbl$pct_variance[keep],
      metric = "Variance explained"
    ),
    data.frame(
      component = var_tbl$component[keep],
      value = var_tbl$cumulative_pct_variance[keep],
      metric = "Cumulative variance explained"
    )
  )

  ggplot(plot_df, aes(x = component, y = value)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    facet_wrap(~ metric, ncol = 1, scales = "free_y") +
    labs(
      title = title,
      x = "Principal component",
      y = "Percent"
    )
}

top_loadings <- function(pca_object, component = 1, n = 8) {
  loadings <- pca_object$rotation[, component]
  ord <- order(abs(loadings), decreasing = TRUE)
  ord <- ord[seq_len(min(n, length(ord)))]
  data.frame(
    component = paste0("PC", component),
    variable = names(loadings)[ord],
    loading = round(loadings[ord], 3),
    absolute_loading = round(abs(loadings[ord]), 3),
    row.names = NULL
  )
}
```

# PCA Demo Using Digits Data

## Load digits data

```{r load-digits, echo = TRUE}
data_file <- "UnsupL_SISBID_2026.Rdata"
if (!file.exists(data_file)) {
  stop("Expected data file not found: ", data_file)
}

load(data_file)

if (!exists("digits")) {
  stop("The data file must contain an object named 'digits'.")
}

if (!exists("imagedigit")) {
  imagedigit <- function(x) {
    side <- sqrt(length(x))
    if (side != floor(side)) {
      stop("Cannot display image: digit vector length is not a perfect square.")
    }
    mat <- matrix(x, nrow = side, byrow = TRUE)
    image(t(apply(mat, 2, rev)), axes = FALSE, col = gray.colors(256))
  }
}

digits <- as.matrix(digits)
digit_labels <- rownames(digits)

if (is.null(digit_labels)) {
  stop("The digits matrix must have row names containing digit labels.")
}

digit_keep <- digit_labels %in% c("3", "8")
dat38 <- digits[digit_keep, , drop = FALSE]
dat38_labels <- factor(digit_labels[digit_keep], levels = c("3", "8"))

cat("Number of selected digit images:", nrow(dat38), "\n")
cat("Number of pixels per image:", ncol(dat38), "\n")
table(dat38_labels)
```

## Visualize a few original images

```{r digit-examples, echo = TRUE, fig.width = 8, fig.height = 3}
example_ids <- c(which(dat38_labels == "3")[1:5], which(dat38_labels == "8")[1:5])
example_ids <- example_ids[!is.na(example_ids)]

old_par <- par(mfrow = c(2, 5), mar = c(0.2, 0.2, 1.5, 0.2))
for (i in example_ids) {
  imagedigit(dat38[i, ])
  title(main = paste("Digit", dat38_labels[i]))
}
par(old_par)
```

## PCA on centered digit pixels

For image data, we center each pixel before PCA. We do not scale pixels here because scaling can over-emphasize nearly constant background pixels. For data sets with variables on very different scales, see the College example below.

```{r digit-pca, echo = TRUE}
digit_pca <- prcomp(dat38, center = TRUE, scale. = FALSE)
digit_var <- variance_table(digit_pca)

head(digit_var, 10)
```

## PC scatterplot

```{r digit-pc-scatter, echo = TRUE}
digit_score_mat <- digit_pca$x[, 1:5, drop = FALSE]
rownames(digit_score_mat) <- NULL
digit_scores <- as.data.frame(digit_score_mat)
digit_scores$digit <- dat38_labels

ggplot(digit_scores, aes(x = PC1, y = PC2, color = digit)) +
  geom_point(alpha = 0.75, size = 2) +
  labs(
    title = "Digits 3 and 8: first two principal components",
    x = paste0("PC1 (", round(digit_var$pct_variance[1], 1), "% variance)"),
    y = paste0("PC2 (", round(digit_var$pct_variance[2], 1), "% variance)"),
    color = "Digit"
  )
```

## Pairs plot for the first five PCs

```{r digit-pairs, echo = TRUE, fig.width = 8, fig.height = 8}
ggpairs(
  digit_scores,
  columns = 1:5,
  mapping = aes(color = digit),
  title = "Pairwise plots of the first five digit PCs"
)
```

## PC Loadings

Each loading image shows a direction of variation in pixel space. Bright and dark regions indicate pixels that move in opposite directions along that component.

```{r digit-loadings, echo = TRUE, fig.width = 8, fig.height = 5}
old_par <- par(mfrow = c(3, 5), mar = c(0.2, 0.2, 1.5, 0.2))
for (i in 1:15) {
  imagedigit(digit_pca$rotation[, i])
  title(main = paste0("PC", i))
}
par(old_par)
```

## Variance explained

```{r digit-variance, echo = TRUE}
plot_variance(digit_var, title = "Digit PCA variance explained", max_components = 30)
```

```{r digit-components-needed, echo = TRUE}
thresholds <- c(50, 75, 90, 95)
components_needed <- sapply(thresholds, function(thresh) {
  which(digit_var$cumulative_pct_variance >= thresh)[1]
})

data.frame(
  cumulative_variance_threshold = paste0(thresholds, "%"),
  components_needed = components_needed
)
```

## Low-rank reconstruction


```{r digit-reconstruction, echo = TRUE, fig.width = 8, fig.height = 3}
reconstruct_digit <- function(row_id, n_components) {
  scores <- digit_pca$x[row_id, 1:n_components, drop = FALSE]
  loadings <- digit_pca$rotation[, 1:n_components, drop = FALSE]
  reconstruction <- scores %*% t(loadings)
  as.numeric(reconstruction + digit_pca$center)
}

row_id <- which(dat38_labels == "3")[1]
component_grid <- c(1, 5, 20, 50)
component_grid <- component_grid[component_grid <= ncol(digit_pca$x)]

old_par <- par(mfrow = c(1, length(component_grid) + 1), mar = c(0.2, 0.2, 1.8, 0.2))
imagedigit(dat38[row_id, ])
title(main = "Original")

for (k in component_grid) {
  imagedigit(reconstruct_digit(row_id, k))
  title(main = paste(k, "PCs"))
}
par(old_par)
```

# PCA via SVD


## Raw SVD without centering

```{r raw-svd, echo = TRUE}
svd_raw <- svd(dat38)
raw_scores <- sweep(svd_raw$u[, 1:5, drop = FALSE], 2, svd_raw$d[1:5], `*`)
colnames(raw_scores) <- paste0("SVD", 1:5)
raw_score_df <- as.data.frame(raw_scores)
raw_score_df$digit <- dat38_labels

raw_var <- data.frame(
  component = seq_along(svd_raw$d),
  pct_variance = 100 * svd_raw$d^2 / sum(svd_raw$d^2)
)
raw_var$cumulative_pct_variance <- cumsum(raw_var$pct_variance)
```

```{r raw-svd-scatter, echo = TRUE}
ggplot(raw_score_df, aes(x = SVD1, y = SVD2, color = digit)) +
  geom_point(alpha = 0.75, size = 2) +
  labs(
    title = "Raw SVD scores for digits 3 and 8",
    x = paste0("SVD1 (", round(raw_var$pct_variance[1], 1), "% raw sum-of-squares)"),
    y = paste0("SVD2 (", round(raw_var$pct_variance[2], 1), "% raw sum-of-squares)"),
    color = "Digit"
  )
```

## SVD Loadings

```{r raw-svd-loadings, echo = TRUE, fig.width = 8, fig.height = 5}
old_par <- par(mfrow = c(3, 5), mar = c(0.2, 0.2, 1.5, 0.2))
for (i in 1:15) {
  imagedigit(svd_raw$v[, i])
  title(main = paste0("SVD", i))
}
par(old_par)
```

## Raw SVD variance explained

```{r raw-svd-variance, echo = TRUE}
plot_variance(raw_var, title = "Raw SVD variance explained", max_components = 30)
```

## Centered SVD and PCA equivalence check

```{r centered-svd-check, echo = TRUE}
centered_digits <- scale(dat38, center = TRUE, scale = FALSE)
svd_centered <- svd(centered_digits)

# Singular vectors are identifiable only up to sign, so compare absolute values.
max_loading_difference <- max(abs(abs(svd_centered$v[, 1]) - abs(digit_pca$rotation[, 1])))
max_loading_difference
```

# PCA Demo Using College Data

## Load College data

```{r load-college, echo = TRUE}
load_college_data <- function() {
  env <- new.env(parent = emptyenv())

  if (requireNamespace("ISLR", quietly = TRUE)) {
    data("College", package = "ISLR", envir = env)
    return(get("College", envir = env))
  }

  if (requireNamespace("ISLR2", quietly = TRUE)) {
    data("College", package = "ISLR2", envir = env)
    return(get("College", envir = env))
  }

  stop("Please install either the ISLR or ISLR2 package to use the College data.")
}

College <- load_college_data()
college_numeric <- vapply(College, is.numeric, logical(1))
cdat <- College[, college_numeric, drop = FALSE]

cat("College data dimensions:", dim(cdat), "\n")
names(cdat)
```

## PCA with centering and scaling

For the College data, centering and scaling are important because variables such as enrollment counts, graduation rates, and dollar amounts are measured on different scales.

```{r college-pca, echo = TRUE}
college_pca <- prcomp(cdat, center = TRUE, scale. = TRUE)
college_var <- variance_table(college_pca)

head(college_var, 10)
```

## PC scatterplot

```{r college-scatter, echo = TRUE}
college_scores <- as.data.frame(college_pca$x[, 1:5, drop = FALSE])
college_scores$college <- rownames(College)
college_scores$Private <- College$Private

ggplot(college_scores, aes(x = PC1, y = PC2, color = Private, label = college)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_text(check_overlap = TRUE, size = 2.5, vjust = 1.2, show.legend = FALSE) +
  labs(
    title = "College data: first two principal components",
    x = paste0("PC1 (", round(college_var$pct_variance[1], 1), "% variance)"),
    y = paste0("PC2 (", round(college_var$pct_variance[2], 1), "% variance)"),
    color = "Private college?"
  )
```

## Pairs plot

```{r college-pairs, echo = TRUE, fig.width = 8, fig.height = 8}
ggpairs(
  college_scores,
  columns = 1:5,
  mapping = aes(color = Private),
  title = "Pairwise plots of the first five College PCs"
)
```

## Loadings: variables that contribute to the patterns

```{r college-loadings-table, echo = TRUE}
college_top_loadings <- rbind(
  top_loadings(college_pca, component = 1, n = 8),
  top_loadings(college_pca, component = 2, n = 8)
)

knitr::kable(college_top_loadings)
```

```{r college-loadings-plot, echo = TRUE, fig.height = 6}
loading_df <- rbind(
  data.frame(
    variable = rownames(college_pca$rotation),
    loading = college_pca$rotation[, 1],
    component = "PC1"
  ),
  data.frame(
    variable = rownames(college_pca$rotation),
    loading = college_pca$rotation[, 2],
    component = "PC2"
  )
)

ggplot(loading_df, aes(x = reorder(variable, loading), y = loading)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ component, ncol = 1, scales = "free_y") +
  labs(
    title = "College PCA loadings for PC1 and PC2",
    x = NULL,
    y = "Loading"
  )
```

## Variance explained

```{r college-variance, echo = TRUE}
plot_variance(college_var, title = "College PCA variance explained", max_components = nrow(college_var))
```

```{r college-components-needed, echo = TRUE}
thresholds <- c(50, 75, 90, 95)
components_needed <- sapply(thresholds, function(thresh) {
  which(college_var$cumulative_pct_variance >= thresh)[1]
})

data.frame(
  cumulative_variance_threshold = paste0(thresholds, "%"),
  components_needed = components_needed
)
```
